---
title: "Ravenswood Peer School Working File"
author: "Qingyang Zhang & Dorna Abdi"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# import data

For more data field references: https://caaspp-elpac.ets.org/caaspp/ResearchFileFormatSB?ps=true&lstTestYear=2022&lstTestType=B

```{r importing_packages}
library(tidyverse)
library(readxl)
library(dplyr)
library(ggplot2)
library(Hmisc) #https://CRAN.R-project.org/package=Hmisc
library(stargazer)
library(RANN) #https://cran.r-project.org/web/packages/RANN/RANN.pdf

library(stats) #hclust
library(FactoMineR)
library(factoextra)
```

```{r loading_data}
caaspp22 <- read.table('data/caaspp.txt',sep='^', header=TRUE)
student_groups <- read_excel('data/lookup.xlsx', sheet = "student_groups")
entities <- read_csv('data/entities.csv')
ravenswood <- read_csv('data/rvsw.csv') 
bh <- read_csv('data/bh.csv') 
```

```{r exporting}
#write.csv(entities,file='data/entities.csv')
#write.csv(caaspp22,file='data/caaspp22.csv')
#ravenswood <- caaspp22 %>% filter(District.Code=='68999')
#write.csv(ravenswood,file='data/rvsw.csv')
#bh <- caaspp22 %>% filter(School.Code=='6044309')
#write.csv(bh,file='data/bh.csv')
```

# Wrangling

```{r full-data}
select_cols <- c('School.Code',
                  'Student.Group.ID',
                  'Students.Tested')

select_groups <- c(1, 128, 31, 170, 75, 76, 74, 77, 78, 79, 80, 144, 4, 90, 121)

caaspp22_ela_pivoted <- caaspp22 %>% 
    # Grade 13 means all grades, Test.ID 1 means SB - English Language Arts/Literacy,
    filter(Grade==13 & Test.ID==1) %>%
    # Type.ID 7 means public schools
    filter(Type.ID==7 | Type.ID==9 | Type.ID==10) %>%
    select(all_of(select_cols)) %>%
    filter(Student.Group.ID %in% select_groups) %>%
    left_join(y=student_groups, by = c("Student.Group.ID" = "Demographic ID")) %>% 
    rename("Demographic.ID.Num" = "Demographic ID Num",
           "Demographic.Name" = "Demographic Name") %>%
    select(all_of(c('School.Code',
                  'Students.Tested',
                  'Demographic.Name'))) %>%
    pivot_wider(names_from = Demographic.Name, values_from = 'Students.Tested')
```

```{r school_type}
caaspp22_grade <- caaspp22 |> 
  group_by(School.Code) |> # 10,258 Unique School Codes
  filter(School.Code != 0 & Grade != 13) |>
  mutate(Grade.Count = n_distinct(Grade)) |>
  mutate(Grade.List = paste0(unique(Grade), collapse = ','))

grade_count <- caaspp22 |> 
  group_by(School.Code) |> # 10,258 Unique School Codes
  filter(School.Code != 0 & Grade != 13) |>
  summarise(Grade.Count = n_distinct(Grade))

grade_list <- caaspp22 |> 
  group_by(School.Code) |> # 10,258 Unique School Codes
  filter(School.Code != 0 & Grade != 13) |>
  summarise(Grade.List = paste0(unique(Grade), collapse = ','))

grades <- grade_count |>
  left_join(y=grade_list, by=c("School.Code" = "School.Code"))

school_type_table <- as.data.frame(
  table(grades$Grade.List, grades$Grade.Count)) |>
  setNames(c('grade_list', 'grade_count', 'freq')) |>
  filter(freq != 0)
```

## Missing data

```{r full-data-missing-val}
caaspp22_ela_filled <- data.frame(caaspp22_ela_pivoted)
caaspp22_ela_filled[caaspp22_ela_filled=='*'] <- '1'
caaspp22_ela_filled[is.na(caaspp22_ela_filled)] <- '0'
caaspp22_ela_filled <- caaspp22_ela_filled %>% mutate_if(is.character, as.numeric)
```

```{r rename-export}
col_names <- names(caaspp22_ela_filled)[3:length(names(caaspp22_ela_filled))]
for (col_name in col_names) {
  name <- paste0('Perc.', col_name)
  caaspp22_ela_filled[name] <- caaspp22_ela_filled[col_name] / caaspp22_ela_filled$`All.Students` * 100
}
caaspp22_ela_demo <- caaspp22_ela_filled %>% select(-col_names)

write.csv(caaspp22_ela_demo,file='data/caaspp22_ela_demo.csv')
```

## Outcome Encoding

```{r scores-distance}
#df by grade
caaspp_ela_outcome <- caaspp22 |>
  select(School.Code, Student.Group.ID, Test.ID, Grade, Students.Enrolled, Students.Tested, Mean.Scale.Score) |>
  filter(School.Code > 0) |>
  filter(Test.ID==1) |>
  filter(Student.Group.ID == 1) |>
  filter(Grade == 3 | Grade == 4 | Grade == 5 | Grade == 6 | Grade == 7 | Grade == 8) |>
  select(School.Code, Grade, Students.Enrolled, Students.Tested, Mean.Scale.Score)

#Grade 3 distance from score (2432)
grade_3 <- caaspp_ela_outcome |>
  filter(Grade == 3) |>
  mutate(distance_from_score_ela = as.numeric(Mean.Scale.Score) - 2432)

#Grade 4 distance from score (2473)
grade_4 <- caaspp_ela_outcome |>
  filter(Grade == 4) |>
  mutate(distance_from_score_ela = as.numeric(Mean.Scale.Score) - 2473)

#Grade 5 distance from score (2502)
grade_5 <- caaspp_ela_outcome |>
  filter(Grade == 5) |>
  mutate(distance_from_score_ela = as.numeric(Mean.Scale.Score) - 2502)

#Grade 6 distance from score (2618)
grade_6 <- caaspp_ela_outcome |>
  filter(Grade == 6) |>
  mutate(distance_from_score_ela = as.numeric(Mean.Scale.Score) - 2618)

#Grade 7 distance from score (2649)
grade_7 <- caaspp_ela_outcome |>
  filter(Grade == 7) |>
  mutate(distance_from_score_ela = as.numeric(Mean.Scale.Score) - 2649)

#Grade 8 distance from score (2668)
grade_8 <- caaspp_ela_outcome |>
  filter(Grade == 8) |>
  mutate(distance_from_score_ela = as.numeric(Mean.Scale.Score) - 2668)

caaspp_ela_score_distance <- rbind(grade_3, grade_4, grade_5, grade_6, grade_7, grade_8) |>
  na.omit()

caaspp_ela_aggregate <- caaspp_ela_score_distance |>
  group_by(School.Code) |>
  summarise_at(vars(distance_from_score_ela), list(name = mean)) |>
  rename("Avg.Score.Dist"="name")

```

```{r merge-output}
caaspp22_ela_score <- caaspp22_ela_demo |> left_join(y=caaspp_ela_aggregate, by = c("School.Code" = "School.Code"))

write.csv(caaspp22_ela_score, file='data/caaspp22_ela_score.csv')
```

# Modeling
Notice that the built-in knn function runs a k-nearest neighbour classification for test set from training set, which does not apply for our situation. 

Potential related: 

*   NN search: https://search.r-project.org/CRAN/refmans/RANN/html/nn2.html
*   NN match: https://kosukeimai.github.io/MatchIt/reference/method_nearest.html
*   KNN regression: https://www.datatechnotes.com/2020/10/knn-regresion-example-in-r.html

```{r importing}
caaspp22_ela <- read_csv('data/caaspp22_ela_demo.csv') %>%
    # drop the column to set School.Code as index
    column_to_rownames(., var = 'School.Code') %>%
    select(-'...1') %>%
    # 223 missing values in Perc%. Potentially bc OG num is 0, then impute 0? drop for now
    drop_na()
```

## Feature Engineering
Before running models, we would like to conduct some feature engineering.

1. Normalize school size (`All.Students`)

```{r preprocessing}
hist(caaspp22_ela$All.Students)
d <- density(caaspp22_ela$All.Students) # returns the density data
plot(d) # plots the results
boxplot(caaspp22_ela$All.Students, data=caaspp22_ela)

entities[entities['School Code']=='1996115', ]

# normalize the school size column
# use log to address outliers 
# add 1e-10 to avoid 0 throwing off the normalization
normalize <- function(x) {
  x <- log(x + 1e-10)
  return ((x - min(x)) / (max(x) - min(x)) * 100) 
}


caaspp22_ela['All.Students'] <- lapply(caaspp22_ela['All.Students'], normalize)
```

Before normalization, we found out that there is an outlier (6421) that is way greater than all other schools (>>3763, the second large). We consider removing the record (City of Angels at Los Angeles Unified).

## Nearest Neighbor

```{r modeling}
nearest <- nn2(caaspp22_ela, caaspp22_ela)
```

```{r qc-outcome}
selectedna <- c('1037456', '115402', '132233')
sample_na <- caaspp22 |> 
  filter(School.Code %in% selectedna) |>
  filter(Test.ID==1) |>
  filter(Student.Group.ID == 1)
```


```{r modeling-loop-all}
ravenswood_school_codes <- caaspp22 %>% 
                                filter(District.Code=='68999') %>% 
                                filter(Type.ID==7 | Type.ID==9 | Type.ID==10) %>%
                                select(School.Code) %>% 
                                unique() %>% 
                                .$School.Code

top10_ravenswood <- data.frame(Rank=numeric(0), 
                               Dist=numeric(0),
                               School.Code=numeric(0))

for (school in ravenswood_school_codes) {
  
  # each ravenswood school has ten rows, i.e. top ten neighbors 
  row_idx <- which(rownames(caaspp22_ela) == as.character(school)) 
  top10_id <- nearest$nn.idx[row_idx,]
  top10_ds <-nearest$nn.dists[row_idx,]
  
  # each row a neighbor with info about rank，dist, school.code
  for (rk in 1:10) {
    nn_idx <- top10_id[rk]
    neighbor_code <- rownames(caaspp22_ela)[nn_idx]
    top10_ravenswood[nrow(top10_ravenswood) + 1, ] <- c(rk, top10_ds[rk], neighbor_code)  
  }
  
}

write.csv(top10_ravenswood, file='data/top10_ravenswood.csv')

```

## Hierarchical Clustering

```{r hclust}
hclusters <- hclust(dist(caaspp22_ela))
plot(hclusters) # uneven distribution over 4 clusters
```

## K-means Clustering
```{r pca}
res.pca <- prcomp(caaspp22_ela, center = TRUE,scale. = TRUE)
print(res.pca)
fviz_eig(res.pca)
fviz_pca_var(res.pca,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )
```


```{r kmeans}
#library(cluster)
library(factoextra)
library(ggpubr)

fviz_nbclust(caaspp22_ela, kmeans, method='wss')
set.seed(1234)
#concern: not exactly same size
km <- kmeans(caaspp22_ela, 4)
fviz_cluster(km, data=caaspp22_ela, geom="point")

```