---
title: "eda"
author: "Qingyang Zhang"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# import data

For more data field references: https://caaspp-elpac.ets.org/caaspp/ResearchFileFormatSB?ps=true&lstTestYear=2022&lstTestType=B

```{r importing_packages}
library(tidyverse)
library(readxl)
library(Hmisc) #https://CRAN.R-project.org/package=Hmisc
library(stargazer)
library(RANN) #https://cran.r-project.org/web/packages/RANN/RANN.pdf
```

```{r loading_data}
caaspp22 <- read.table('data/caaspp.txt',sep='^', header=TRUE)
student_groups <- read_excel('data/lookup.xlsx', sheet = "student_groups")
entities <- read_csv('data/entities.csv')
ravenswood <- read_csv('data/rvsw.csv') 
bh <- read_csv('data/bh.csv') 
```

```{r exporting}
#write.csv(entities,file='data/entities.csv')
#write.csv(caaspp22,file='data/caaspp22.csv')
#ravenswood <- caaspp22 %>% filter(District.Code=='68999')
#write.csv(ravenswood,file='data/rvsw.csv')
#bh <- caaspp22 %>% filter(School.Code=='6044309')
#write.csv(bh,file='data/bh.csv')
```


# EDA on Belle Haven 

---
We sample the records of the Belle Haven Elementary School ELA test results as the toy data set to explore for data wrangling. 

🛑 Decision To Be Done:
* what variable to choose as the number of students when calculating the proportion of student composition for each student subgroup (consider that there could be slight difference across subjects for tested-related variables) 🅰  use `students.tested`
* how to handle missing values for certain student subgroup use? 
* to what level of granularity of student subgroup we are going to use 🅰 ignore two-level first

```{r}
select_cols <- c('Student.Group.ID',
                      'Total.Tested.at.Reporting.Level', 
                      'Total.Tested.with.Scores.at.Reporting.Level', 
                      'Students.Enrolled', 
                      'Students.Tested')
bh_allgrades <- bh %>% 
    # Grade 13 means all grades, Test.ID 1 means SB - English Language Arts/Literacy
    filter(Grade==13 & Test.ID==1) %>%
    select(all_of(select_cols)) %>%
    left_join(y=student_groups, by = c("Student.Group.ID" = "Demographic ID")) %>% 
    rename("Demographic.ID.Num" = "Demographic ID Num",
           "Demographic.Name" = "Demographic Name",
           "Student.Group" = "Student Group")
```

##### Assume using 'Students.Tested'
With respect to what variable to use as the number of students when calculating the proportion of student composition for each student subgroup, for simplicity, we use `Students.Tested` for EDA and detailed illustrations. We observe that 
* there are 12 ways of describing subgroups, and in total 45 subgroups
* there are missing data, marked as `*`, which seems to differ from 0. e.g.
  
  0	|American Indian or Alaska Native	| Race and Ethnicity

  \* |	Filipino	| Race and Ethnicity
  
```{r}
bh_allgrades_a <- bh_allgrades %>%
    select(c('Students.Tested', 'Demographic.ID.Num', 'Demographic.Name', 'Student.Group'))
bh_allgrades_a %>%
    count(Student.Group) %>%
    pivot_wider(names_from = Student.Group, values_from = n)
```


##### Assume replace missing value with 0
```{r }
bh_allgrades_a <- bh_allgrades_a %>%
    # notice NAs will be introduced by coercion
    mutate_at('Students.Tested', as.numeric) %>%
    # replace NA with 0
    replace_na(list(Students.Tested = 0)) 
```

```{r}
bh_allgrades_a <- bh_allgrades_a %>% 
    # create a new column for percentage 
    mutate(Perc = Students.Tested / max(bh_allgrades_a$Students.Tested)*100) %>%
    mutate(Student.Group.Desc = paste(Demographic.ID.Num, Demographic.Name, Student.Group, sep = '_'))
```

```{r}
bh_allgrades_a %>% 
    select(c('Student.Group.Desc', 'Perc')) %>%
    pivot_wider(names_from = Student.Group.Desc, values_from = Perc)
```

### on district data

```{r ravenswood}

select_cols <- c('School.Code',
                  'Student.Group.ID',
                  'Students.Tested')

select_groups <- c(1, 128, 31, 170, 75, 76, 74, 77, 78, 79, 80, 144, 4, 90, 121)

ravenswood_s <- ravenswood %>% 
    # Grade 13 means all grades, Test.ID 1 means SB - English Language Arts/Literacy
    filter(Grade==13 & Test.ID==1) %>%
    select(all_of(select_cols)) %>%
    filter(Student.Group.ID %in% select_groups) %>%
    left_join(y=student_groups, by = c("Student.Group.ID" = "Demographic ID")) %>% 
    rename("Demographic.ID.Num" = "Demographic ID Num",
           "Demographic.Name" = "Demographic Name") %>%
    select(all_of(c('School.Code',
                  'Students.Tested',
                  'Demographic.Name'))) %>%
    pivot_wider(names_from = School.Code, values_from = 'Students.Tested')
                  #'Demographic.Name'))
ravenswood_s

#latexTabular(format.df(ravenswood_s))
#write.csv(ravenswood_s ,file='data/rvsw_agg.csv')
```

# Wrangling

```{r full-data}
caaspp22_ela_pivoted <- caaspp22 %>% 
    # Grade 13 means all grades, Test.ID 1 means SB - English Language Arts/Literacy,
    filter(Grade==13 & Test.ID==1) %>%
    # Type.ID 7 means public schools
    filter(Type.ID==7 | Type.ID==9 | Type.ID==10) %>%
    select(all_of(select_cols)) %>%
    filter(Student.Group.ID %in% select_groups) %>%
    left_join(y=student_groups, by = c("Student.Group.ID" = "Demographic ID")) %>% 
    rename("Demographic.ID.Num" = "Demographic ID Num",
           "Demographic.Name" = "Demographic Name") %>%
    select(all_of(c('School.Code',
                  'Students.Tested',
                  'Demographic.Name'))) %>%
    pivot_wider(names_from = Demographic.Name, values_from = 'Students.Tested')
```


```{r full-data-missing-val}
caaspp22_ela_filled <- data.frame(caaspp22_ela_pivoted)
caaspp22_ela_filled[caaspp22_ela_filled=='*'] <- '1'
caaspp22_ela_filled[is.na(caaspp22_ela_filled)] <- '0'
caaspp22_ela_filled <- caaspp22_ela_filled %>% mutate_if(is.character, as.numeric)
```

```{r rename-export}
col_names <- names(caaspp22_ela_filled)[3:length(names(caaspp22_ela_filled))]
for (col_name in col_names) {
  name <- paste0('Perc.', col_name)
  caaspp22_ela_filled[name] <- caaspp22_ela_filled[col_name] / caaspp22_ela_filled$`All.Students` * 100
}
caaspp22_ela_demo <- caaspp22_ela_filled %>% select(-col_names)

write.csv(caaspp22_ela_demo,file='data/caaspp22_ela_demo.csv')
```

```{r scores-distance}
#df by grade
caaspp_ela_outcome <- caaspp22 |>
  select(School.Code, Student.Group.ID, Test.ID, Grade, Students.Enrolled, Students.Tested, Mean.Scale.Score) |>
  filter(School.Code > 0) |>
  filter(Test.ID==1) |>
  filter(Student.Group.ID == 1) |>
  filter(Grade == 3 | Grade == 4 | Grade == 5 | Grade == 6 | Grade == 7 | Grade == 8) |>
  select(School.Code, Grade, Students.Enrolled, Students.Tested, Mean.Scale.Score)

#Grade 3 distance from score (2432)
grade_3 <- caaspp_ela_outcome |>
  filter(Grade == 3) |>
  mutate(distance_from_score_ela = as.numeric(Mean.Scale.Score) - 2432)

#Grade 4 distance from score (2473)
grade_4 <- caaspp_ela_outcome |>
  filter(Grade == 4) |>
  mutate(distance_from_score_ela = as.numeric(Mean.Scale.Score) - 2473)

#Grade 5 distance from score (2502)
grade_5 <- caaspp_ela_outcome |>
  filter(Grade == 5) |>
  mutate(distance_from_score_ela = as.numeric(Mean.Scale.Score) - 2502)

#Grade 6 distance from score (2618)
grade_6 <- caaspp_ela_outcome |>
  filter(Grade == 6) |>
  mutate(distance_from_score_ela = as.numeric(Mean.Scale.Score) - 2618)

#Grade 7 distance from score (2649)
grade_7 <- caaspp_ela_outcome |>
  filter(Grade == 7) |>
  mutate(distance_from_score_ela = as.numeric(Mean.Scale.Score) - 2649)

#Grade 8 distance from score (2668)
grade_8 <- caaspp_ela_outcome |>
  filter(Grade == 8) |>
  mutate(distance_from_score_ela = as.numeric(Mean.Scale.Score) - 2668)

caaspp_ela_score_distance <- rbind(grade_3, grade_4, grade_5, grade_6, grade_7, grade_8) |>
  na.omit()

caaspp_ela_aggregate <- caaspp_ela_score_distance |>
  group_by(School.Code) |>
  summarise_at(vars(distance_from_score_ela), list(name = mean)) |>
  rename("Avg.Score.Dist"="name")

```

```{r merge-output}
caaspp22_ela_score <- caaspp22_ela_demo |> left_join(y=caaspp_ela_aggregate, by = c("School.Code" = "School.Code"))

write.csv(caaspp22_ela_score, file='data/caaspp22_ela_score.csv')
```


# KNN
Notice that the built-in knn function runs a k-nearest neighbour classification for test set from training set, which does not apply for our situation. 

Potential related: 

*   NN search: https://search.r-project.org/CRAN/refmans/RANN/html/nn2.html
*   NN match: https://kosukeimai.github.io/MatchIt/reference/method_nearest.html
*   KNN regression: https://www.datatechnotes.com/2020/10/knn-regresion-example-in-r.html

```{r modeling}
caaspp22_ela <- read_csv('data/caaspp22_ela_demo.csv') %>%
    # drop the column to set School.Code as index
    column_to_rownames(., var = 'School.Code') %>%
    select(-'...1') %>%
    # 223 missing values in Perc%. Potentially bc OG num is 0, then impute 0? drop for now
    drop_na()
                
nearest <- nn2(caaspp22_ela, caaspp22_ela)

```


```{r modeling-bh}

caaspp22_ela['6044309', ]
row_idx <- which(rownames(caaspp22_ela) == "6044309") 
top10bh <- nearest$nn.idx[row_idx,]
top10bhdis <-nearest$nn.dists[row_idx,]

```

```{r modeling-loop-all}
ravenswood_school_codes <- caaspp22 %>% 
                                filter(District.Code=='68999') %>% 
                                filter(Type.ID==7 | Type.ID==9 | Type.ID==10) %>%
                                select(School.Code) %>% 
                                unique() %>% 
                                .$School.Code

top10_ravenswood <- data.frame(Rank=numeric(0), 
                               Dist=numeric(0),
                               School.Code=numeric(0))

for (school in ravenswood_school_codes) {
  
  # each ravenswood school has ten rows, i.e. top ten neighbors 
  row_idx <- which(rownames(caaspp22_ela) == as.character(school)) 
  top10_id <- nearest$nn.idx[row_idx,]
  top10_ds <-nearest$nn.dists[row_idx,]
  
  # each row a neighbor with info about rank，dist, school.code
  for (rk in 1:10) {
    nn_idx <- top10_id[rk]
    neighbor_code <- rownames(caaspp22_ela)[nn_idx]
    top10_ravenswood[nrow(top10_ravenswood) + 1, ] <- c(rk, top10_ds[rk], neighbor_code)  
  }
  
}

write.csv(top10_ravenswood, file='data/top10_ravenswood.csv')

```